\documentclass[11pt]{article}

% Standard packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{natbib}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,positioning,fit,calc,decorations.pathreplacing}

\geometry{margin=1in}

% Theorem environments
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}

% Custom commands
\newcommand{\efa}{\textsc{EFA}}
\newcommand{\cmpg}{\textsc{CMPG}}
\newcommand{\fwrl}{\textsc{FWRL}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\calG}{\mathcal{G}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calD}{\mathcal{D}} % Criteria space (distinct from generator \calC)

\title{\textbf{Evaluation-First Attention: Specification-Driven Generation \\ via Dynamic Rubric Conditioning and Failure-Weighted Reattention}}

\author{
Karthick Raja Mohan \\
Independent Researcher \\
Chennai, India \\
\texttt{karthickrajam18@gmail.com}
}

\date{March 2026}

\begin{document}

\maketitle

% ============================================================================
% ABSTRACT
% ============================================================================
\begin{abstract}
Current large language model (LLM) generation pipelines treat evaluation as a post-hoc measurement step: generate first, evaluate later. This separation means models generate ``blindly'' without explicit awareness of the quality dimensions they must satisfy, and refinement feedback---when provided---is holistic rather than dimension-specific. We propose \textbf{Evaluation-First Attention} (\efa{}), an architecture that inverts this paradigm by generating query-specific evaluation criteria \emph{before} generation and using them as structured conditioning targets throughout the generation process. \efa{} introduces two novel mechanisms: (1) \textbf{Criteria-Masked Progressive Generation} (\cmpg{}), which progressively unmasks quality dimensions during generation analogous to causal token masking but over evaluation criteria, and (2) \textbf{Failure-Weighted Reattention Loop} (\fwrl{}), which maps per-criterion evaluation scores to criterion emphasis adjustments, amplifying focus on failing dimensions in subsequent generation passes. We formalize the architecture, analyze convergence properties under monotonic improvement assumptions, and present an experimental framework with six baselines and four ablation conditions. To our knowledge, no prior work in the general text generation domain uses dynamically-generated evaluation rubrics as inference-time generation specifications with failure-proportional reweighting.
\end{abstract}

% ============================================================================
% 1. INTRODUCTION
% ============================================================================
\section{Introduction}

The dominant paradigm for improving LLM output quality follows a \emph{generate-then-evaluate} pattern: a model produces an output, a separate evaluation step scores it, and---optionally---the model refines based on feedback \citep{madaan2023selfrefine, shinn2023reflexion, gou2024critic}. While effective, this paradigm suffers from three fundamental limitations.

First, \textbf{evaluation is disconnected from generation}. The model has no explicit awareness of what quality dimensions it must satisfy during the generation process itself. Rubric-based reward models \citep{li2025openrubrics, zhang2025carmo} generate sophisticated per-query evaluation criteria but use them exclusively for post-hoc scoring, never as generation targets. This is analogous to writing code without knowing the test suite---possible, but needlessly inefficient.

Second, \textbf{refinement feedback is holistic, not dimensional}. Self-Refine \citep{madaan2023selfrefine} produces textual feedback like ``the response lacks specificity,'' requiring the model to implicitly parse which quality dimensions failed. The self-refinement literature itself identifies this as the core bottleneck: diagnosis---not repair---is the primary challenge \citep{lee2025refinebench}. Models achieve near-perfect refinement when given targeted, structured feedback but struggle with self-diagnosis.

Third, \textbf{all quality dimensions receive equal emphasis}. Whether a response perfectly satisfies relevance but catastrophically fails on factual accuracy, subsequent refinement passes treat all dimensions uniformly. No existing system allocates more generation ``budget'' to failing dimensions.

We propose \textbf{Evaluation-First Attention} (\efa{}), an architecture that addresses all three limitations by inverting the generation-evaluation relationship. \efa{} operates in three phases:

\begin{enumerate}[leftmargin=*]
    \item \textbf{Criteria Generation}: Before any generation, a dedicated LLM call analyzes the user prompt and produces an ordered set of query-specific evaluation criteria, each with a measurable scoring rubric.
    \item \textbf{Criteria-Masked Progressive Generation}: A single generator LLM produces the response conditioned on criteria that are progressively unmasked---the generator first sees criterion $c_1$ alone, then $\{c_1, c_2\}$, and so on---analogous to causal masking over quality dimensions rather than tokens.
    \item \textbf{Failure-Weighted Reattention}: The output is evaluated per-criterion. Failed criteria receive mathematically boosted emphasis weights in subsequent generation passes, creating a convergent refinement loop that surgically targets weak dimensions.
\end{enumerate}

The key insight is conceptually simple: \emph{what if LLMs knew what ``good'' looks like before they started writing?} By generating evaluation criteria first and using them as structured generation specifications with failure-aware reweighting, \efa{} transforms evaluation from a passive measurement tool into an active generation target.

Our contributions are:
\begin{enumerate}[leftmargin=*]
    \item We formalize the \textbf{Evaluation-First Attention} architecture, the first system that uses dynamically-generated evaluation criteria as inference-time generation specifications with failure-weighted reattention (\S\ref{sec:method}).
    \item We introduce \textbf{Criteria-Masked Progressive Generation} (\cmpg{}), a novel conditioning mechanism over quality dimensions with no prior art in any generation paradigm (\S\ref{sec:cmpg}).
    \item We introduce the \textbf{Failure-Weighted Reattention Loop} (\fwrl{}), which maps per-criterion evaluation failures to emphasis weight modifications---analogous to focal loss \citep{lin2017focal} but applied to evaluation dimensions at inference time (\S\ref{sec:fwrl}).
    \item We analyze convergence properties, provide computational complexity bounds, and design a rigorous experimental framework with six baselines and four ablation conditions (\S\ref{sec:convergence}, \S\ref{sec:experiments}).
\end{enumerate}

% ============================================================================
% 2. RELATED WORK
% ============================================================================
\section{Related Work}

\subsection{Rubric Generation and Criteria-Based Evaluation}

Recent work has demonstrated the value of dynamic, query-specific evaluation criteria. CARMO \citep{zhang2025carmo} generates context-aware criteria for reward modeling. OpenRubrics \citep{li2025openrubrics} and RRD \citep{wang2026rrd} produce and refine rubrics for preference prediction. G-Eval \citep{liu2023geval} uses chain-of-thought to generate evaluation steps before scoring. AutoSCORE \citep{chen2025autoscore} employs multi-agent rubric-aligned scoring.

\textbf{Critical gap}: All existing rubric systems use criteria exclusively for evaluation or reward modeling---\emph{never as generation specifications}. \efa{} closes this gap by feeding criteria into the generation process as structured conditioning targets.

\subsection{Iterative Self-Refinement}

Self-Refine \citep{madaan2023selfrefine} pioneered iterative LLM self-improvement through generate-feedback-refine loops. Reflexion \citep{shinn2023reflexion} adds verbal reinforcement learning with episodic memory. CRITIC \citep{gou2024critic} incorporates tool-interactive critique. MAgICoRe \citep{chen2025magicore} uses multi-agent coarse-to-fine refinement with process reward models.

\textbf{Critical gap}: All refinement systems provide holistic or textual feedback. None decompose failure into per-criterion scores mapped to generation emphasis adjustments. \efa{}'s failure-weighted reattention provides structured, numerical, dimension-specific feedback that directly modifies criterion emphasis in subsequent passes.

\subsection{Controlled and Constrained Generation}

Controllable text generation approaches steer outputs toward desired attributes. FUDGE \citep{yang2021fudge} uses future discriminators. PPLM \citep{dathathri2020plug} perturbs hidden representations. Classifier-free guidance \citep{ho2022classifier} interpolates conditional and unconditional scores. Constitutional AI \citep{bai2022constitutional} uses fixed principles to guide critique-revision during training data synthesis, but these principles are pre-defined (not query-specific) and applied during training, not at inference time. Recent multi-attribute control methods handle multiple constraints simultaneously.

\textbf{Critical gap}: These methods operate at the token level with fixed, pre-defined attributes. \efa{} operates at the rubric level with dynamically-generated, query-specific criteria and progressive unmasking---a fundamentally different granularity.

\subsection{Multi-Agent and Ensemble Generation}

LLM-Blender \citep{jiang2023llmblender} ensembles outputs via pairwise ranking and generative fusion. FusioN \citep{agarwal2025fusion} synthesizes multiple candidates into superior outputs. Multi-Agent Debate (MAD) frameworks enable agents to collaboratively refine answers through discussion rounds.

\textbf{Critical gap}: Multi-agent systems generate without criteria awareness, then select or synthesize post-hoc. \efa{} conditions generation on explicit criteria from the start, reducing dependence on post-hoc selection and synthesis.

\subsection{Test-Driven Development for Code Generation}

In the code domain, TDD-inspired approaches generate test cases before code \citep{ravi2025llmloop}. Property-based testing bridges generation and validation by verifying generated code against specifications.

\textbf{Critical gap}: Limited to code with executable tests (binary pass/fail). \efa{} generalizes specification-driven generation to arbitrary text quality dimensions with continuous scoring and emphasis-level feedback.

% ============================================================================
% 3. METHOD
% ============================================================================
\section{Evaluation-First Attention Architecture}
\label{sec:method}

\subsection{Overview}

Given a user prompt $P$, \efa{} produces a response $R^*$ through three formally-defined components operating in a closed loop (Figure~\ref{fig:architecture}):

\begin{enumerate}[leftmargin=*]
    \item \textbf{Criteria Generator} $\calC$: Analyzes $P$ and produces an ordered criteria set $C = \{c_1, c_2, \ldots, c_n\}$ with initial emphasis weights $\mathbf{w}^{(0)} \in \R^n$.
    \item \textbf{Criteria-Masked Progressive Generator} $\calG$: Generates response $R^{(k)}$ at iteration $k$ with criteria-conditioned prompting under progressive masking.
    \item \textbf{Failure-Weighted Evaluator} $\calE$: Scores $R^{(k)}$ per-criterion, producing scores $\mathbf{s}^{(k)} \in [0,1]^n$, and computes updated weights $\mathbf{w}^{(k+1)}$ that amplify emphasis on failing dimensions.
\end{enumerate}

The system iterates until $\min(\mathbf{s}^{(k)}) \geq \tau$ (all criteria pass threshold $\tau$) or $k = K_{\max}$.

\textbf{Single-model design}: \efa{} uses a single LLM throughout (the same model serves as criteria generator, progressive generator, and evaluator via different prompts). This is a deliberate design choice: it eliminates inter-model coordination overhead and isolates the architectural contribution from model selection effects. However, the framework naturally extends to heterogeneous multi-model configurations where different models handle different components.

% ============================================================================
% ARCHITECTURE FIGURE
% ============================================================================
\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=1.0cm and 1.5cm,
    >={Stealth[length=3mm]},
    box/.style={draw, rounded corners, minimum width=3.2cm, minimum height=0.9cm, align=center, font=\small},
    critbox/.style={box, fill=blue!12, draw=blue!50},
    genbox/.style={box, fill=green!12, draw=green!50},
    evalbox/.style={box, fill=red!12, draw=red!50},
    iobox/.style={box, fill=gray!12, draw=gray!50},
    arrow/.style={->, thick},
    dasharrow/.style={->, thick, dashed, red!60},
]

% Input
\node[iobox] (input) {User Prompt $P$};

% Criteria Generator
\node[critbox, below=0.8cm of input] (cg) {\textbf{Criteria Generator} $\calC$};
\node[right=0.3cm of cg, font=\scriptsize\itshape, text=blue!70] (cglabel) {Generates $C = \{c_1, \ldots, c_n\}$};

% CMPG Block
\node[genbox, below=0.8cm of cg, minimum width=5.5cm, minimum height=2.4cm] (cmpgbox) {};
\node[above=0.05cm of cmpgbox.north, font=\small\bfseries, text=green!50!black] {Criteria-Masked Progressive Generation ($\calG$)};

% Sub-steps inside CMPG
\node[draw, rounded corners, fill=green!5, minimum width=4.8cm, minimum height=0.5cm, font=\scriptsize, align=center] at ([yshift=-0.15cm]cmpgbox.center) (sub) {
    $d_1^{(k)} \xrightarrow{\{c_1\}}  d_2^{(k)} \xrightarrow{\{c_1,c_2\}} \cdots \xrightarrow{\{c_1,\ldots,c_n\}} R^{(k)}$
};
\node[below=0.05cm of sub, font=\scriptsize\itshape, text=green!40!black] {Progressive unmasking with emphasis weights $\mathbf{w}^{(k)}$};

% Evaluator
\node[evalbox, below=0.8cm of cmpgbox, minimum width=5.5cm] (eval) {\textbf{Failure-Weighted Evaluator} $\calE$};
\node[below=0.05cm of eval, font=\scriptsize\itshape, text=red!50!black] (evallabel) {Scores: $\mathbf{s}^{(k)} = (s_1, \ldots, s_n)$ \quad Reweights: $\mathbf{w}^{(k+1)}$};

% Decision
\node[draw, diamond, aspect=2.5, below=1.0cm of evallabel, fill=yellow!15, draw=yellow!60, font=\small, align=center, inner sep=1pt] (decide) {$\min(\mathbf{s}^{(k)}) \geq \tau$\,?};

% Output
\node[iobox, below=0.8cm of decide] (output) {Final Output $R^*$};

% Arrows
\draw[arrow] (input) -- (cg);
\draw[arrow] (cg) -- node[right, font=\scriptsize] {$C, \mathbf{w}^{(0)}$} (cmpgbox.north);
\draw[arrow] (cmpgbox.south) -- node[right, font=\scriptsize] {$R^{(k)}$} (eval);
\draw[arrow] (eval) -- (decide);
\draw[arrow] (decide) -- node[right, font=\scriptsize] {Yes} (output);

% Feedback loop
\draw[dasharrow] (decide.west) -- node[above, font=\scriptsize, text=red!60] {No} ++(-2.5,0) |- node[left, font=\scriptsize, text=red!60, near start] {$\mathbf{w}^{(k+1)}$ (boosted)} (cmpgbox.west);

% Checkpoint annotation
\node[right=0.3cm of decide, font=\scriptsize\itshape, text=gray] {+ criterion checkpoint};

\end{tikzpicture}
\caption{\textbf{The Evaluation-First Attention (\efa{}) architecture.} A Criteria Generator produces query-specific evaluation criteria before generation begins. The Criteria-Masked Progressive Generator produces a response by incrementally unmasking criteria (analogous to causal token masking but over quality dimensions). The Failure-Weighted Evaluator scores each criterion independently and computes boosted emphasis weights for failing dimensions. The system loops (dashed red arrow) until all criteria pass threshold $\tau$ or maximum iterations are reached. A criterion checkpoint mechanism prevents regression on previously-passing criteria.}
\label{fig:architecture}
\end{figure}

\subsection{Component 1: Criteria Generator ($\calC$)}

\begin{definition}[Criteria Generator]
Given prompt $P$, the criteria generator $\calC: \mathcal{P} \rightarrow (\calD^n, \R^n)$ produces:
\begin{equation}
    \calC(P) = \left( \{c_1, c_2, \ldots, c_n\}, \; \mathbf{w}^{(0)} \right)
\end{equation}
where each criterion $c_i = (\text{name}_i, \text{definition}_i, \text{rubric}_i)$ consists of a dimension name, a natural language definition, and a scoring rubric mapping scores $\{1, 2, 3, 4, 5\}$ to quality descriptions.
\end{definition}

Criteria are generated dynamically per-query, not from a fixed taxonomy. The generator LLM is prompted to produce non-overlapping, measurable dimensions relevant to the specific prompt. Initial weights are uniform: $w_i^{(0)} = 1/n$ for all $i$.

\textbf{Criteria ordering}: The generator produces criteria in descending order of fundamentality. For example, factual accuracy precedes stylistic polish, as satisfying later criteria is meaningless if fundamental criteria fail. This ordering determines the unmasking sequence in \cmpg{}.

\subsection{Component 2: Criteria-Masked Progressive Generation ($\calG$)}
\label{sec:cmpg}

This is \efa{}'s first novel mechanism. We introduce masking over quality dimensions, analogous to causal masking over token positions in autoregressive transformers.

\begin{definition}[Criteria Mask]
At generation sub-step $j \in \{1, \ldots, n\}$, the criteria mask $M_j \in \{0, 1\}^n$ is defined as:
\begin{equation}
    M_j[i] = \begin{cases} 1 & \text{if } i \leq j \\ 0 & \text{if } i > j \end{cases}
\end{equation}
\end{definition}

\textbf{Progressive generation process}: Rather than presenting all criteria simultaneously, \cmpg{} operates in $n$ sub-steps within each iteration $k$:

\begin{enumerate}[leftmargin=*]
    \item \textbf{Sub-step 1}: Generate draft $d_1^{(k)}$ conditioned on $\{c_1\}$ only (criteria $c_2, \ldots, c_n$ are masked). The generator focuses entirely on the most fundamental quality dimension.
    \item \textbf{Sub-step 2}: Refine $d_1^{(k)}$ into $d_2^{(k)}$ conditioned on $\{c_1, c_2\}$. The generator adjusts the draft to additionally satisfy $c_2$ while maintaining $c_1$.
    \item \textbf{Sub-step $j$}: Refine $d_{j-1}^{(k)}$ into $d_j^{(k)}$ conditioned on $\{c_1, \ldots, c_j\}$ with emphasis weights $\mathbf{w}^{(k)} \odot M_j$.
    \item \textbf{Sub-step $n$}: Final draft $d_n^{(k)} = R^{(k)}$ is conditioned on all criteria.
\end{enumerate}

Formally, at sub-step $j$ of iteration $k$:
\begin{equation}
    d_j^{(k)} = \calG\left(P, \; d_{j-1}^{(k)}, \; \{c_i : M_j[i] = 1\}, \; \mathbf{w}^{(k)} \odot M_j \right)
\end{equation}
where $d_0^{(k)}$ is empty for $k=1$ or the previous iteration's output $R^{(k-1)}$ for $k > 1$.

\textbf{Weight incorporation mechanism}: The emphasis weights $\mathbf{w}^{(k)}$ are incorporated into the generator prompt through \emph{weighted criterion representation}. At each sub-step, the prompt to $\calG$ includes:
\begin{enumerate}[leftmargin=*]
    \item The user prompt $P$.
    \item The current draft $d_{j-1}^{(k)}$ (if available).
    \item Each visible criterion $c_i$ (where $M_j[i] = 1$) annotated with its emphasis level, expressed as a natural language priority directive:
\end{enumerate}

\begin{equation}
    \text{prompt}_j^{(k)} = P \;\|\; d_{j-1}^{(k)} \;\|\; \bigoplus_{i : M_j[i]=1} \left[ \text{priority}(\hat{w}_i^{(k)}) \;\|\; c_i \right]
\label{eq:prompt_construction}
\end{equation}

where $\|$ denotes prompt concatenation, $\oplus$ denotes ordered assembly of criterion blocks, and $\text{priority}(\hat{w}_i^{(k)})$ maps the normalized weight to a natural language emphasis directive:

\begin{equation}
    \text{priority}(w) = \begin{cases}
    \texttt{[CRITICAL---highest priority]} & \text{if } w > 2/n \\
    \texttt{[HIGH---elevated priority]} & \text{if } w > 1.5/n \\
    \texttt{[STANDARD---normal priority]} & \text{if } w \geq 0.5/n \\
    \texttt{[SECONDARY---lower priority]} & \text{if } w < 0.5/n
    \end{cases}
\label{eq:priority}
\end{equation}

Note that all thresholds are defined relative to the uniform weight $1/n$, ensuring consistent behavior regardless of the number of criteria. A criterion receives \texttt{CRITICAL} priority only when its weight exceeds twice the uniform baseline, which occurs after substantial failure-weighted boosting.

This design choice avoids modifying model internals (attention heads, hidden states) and instead operates entirely through prompt engineering, making \efa{} applicable to any LLM accessible via API. Prior work on instruction following has demonstrated that LLMs are sensitive to emphasis cues and priority ordering within prompts, supporting the effectiveness of structured priority directives for steering generation behavior.

\textbf{Intuition}: This progressive unmasking ensures foundational quality dimensions are satisfied before the generator attempts to optimize for secondary dimensions. It prevents the common failure mode where models optimize for surface-level quality (fluency, style) at the expense of fundamental correctness.

\subsection{Component 3: Failure-Weighted Reattention Loop ($\calE$)}
\label{sec:fwrl}

This is \efa{}'s second novel mechanism. After generating $R^{(k)}$, the evaluator scores it per-criterion and computes reattention weights.

\begin{definition}[Per-Criterion Evaluation]
The evaluator $\calE$ scores the response against each criterion independently:
\begin{equation}
    \mathbf{s}^{(k)} = \calE(R^{(k)}, C) = \left( s_1^{(k)}, s_2^{(k)}, \ldots, s_n^{(k)} \right) \in [0, 1]^n
\end{equation}
where $s_i^{(k)} = \text{score}(R^{(k)}, c_i) / 5$ is the normalized rubric score for criterion $c_i$, obtained by prompting the LLM with the response $R^{(k)}$, criterion $c_i$, and its associated rubric.
\end{definition}

\begin{definition}[Failure-Weighted Reattention]
The emphasis weight for criterion $c_i$ at iteration $k+1$ is updated as:
\begin{equation}
    w_i^{(k+1)} = w_i^{(k)} \cdot \left(1 + \alpha \cdot \max\left(0, \; \tau - s_i^{(k)}\right)\right)
\label{eq:reattention}
\end{equation}
where $\tau \in (0, 1]$ is the passing threshold and $\alpha > 0$ is the reattention strength hyperparameter. Weights are then renormalized: $\hat{w}_i^{(k+1)} = w_i^{(k+1)} / \sum_j w_j^{(k+1)}$.
\end{definition}

\textbf{Properties of Equation~\ref{eq:reattention}}:
\begin{itemize}[leftmargin=*]
    \item \textbf{Passed criteria}: If $s_i^{(k)} \geq \tau$, then $\max(0, \tau - s_i^{(k)}) = 0$ and $w_i^{(k+1)} = w_i^{(k)}$. Weights remain unchanged for satisfied dimensions.
    \item \textbf{Failed criteria}: If $s_i^{(k)} < \tau$, the weight increases proportionally to the severity of failure. A score of $0.2$ with $\tau = 0.6$ produces a larger boost ($\alpha \cdot 0.4$) than a score of $0.5$ ($\alpha \cdot 0.1$).
    \item \textbf{Focal property}: This is analogous to focal loss \citep{lin2017focal}, which down-weights well-classified examples and focuses training on hard negatives. \fwrl{} applies this principle to quality dimensions at inference time rather than training time.
    \item \textbf{Renormalization}: Normalizing to $\sum_i \hat{w}_i = 1$ ensures that boosting failing criteria \emph{relatively} de-emphasizes passing ones, creating a zero-sum reallocation of generation emphasis.
\end{itemize}

\textbf{Criterion checkpoint mechanism}: To prevent regression on previously-passing criteria, we implement a safeguard: if $s_i^{(k+1)} < s_i^{(k)} - \epsilon$ for a criterion that satisfied $s_i^{(k)} \geq \tau$, the system retains the best-observed score for that criterion and marks it as ``locked.'' Locked criteria retain their peak score for convergence tracking and are excluded from further reweighting, ensuring that the reattention mechanism does not destabilize already-satisfactory dimensions. Note that the full response text is not decomposed per-criterion; rather, the locking mechanism prevents the reweighting loop from over-correcting at the expense of previously-passing dimensions. This mechanism is formally integrated into Algorithm~\ref{alg:efa}.

\subsection{Convergence Analysis}
\label{sec:convergence}

We analyze the convergence properties of \efa{} under stated assumptions. We present this as a proposition rather than a theorem, as the key assumption---that LLMs improve on focused criteria---is empirical rather than provable.

\begin{proposition}[Convergence under Focused Improvement]
\label{prop:convergence}
Assume the following two properties hold:
\begin{enumerate}
    \item \textbf{Focused Improvement}: When criterion $c_i$ receives elevated emphasis weight ($\hat{w}_i > 1/n$), the LLM produces output with $s_i^{(k+1)} \geq s_i^{(k)} + \delta$ for some minimum improvement $\delta > 0$.
    \item \textbf{Bounded Regression}: For any criterion $c_i$ with $s_i^{(k)} \geq \tau$, re-generation satisfies $s_i^{(k+1)} \geq \tau - \epsilon$ for regression tolerance $\epsilon < \delta$.
\end{enumerate}
Then the number of failing criteria $|\{i : s_i^{(k)} < \tau\}|$ is non-increasing across iterations, and the system converges (all criteria passing or locked) in at most $K \leq n \cdot \lceil \tau / \delta \rceil$ iterations.
\end{proposition}

\begin{proof}[Proof sketch]
By Bounded Regression, once a criterion passes threshold $\tau$, it either remains above $\tau - \epsilon$ or is locked by the checkpoint mechanism at its passing score. By Focused Improvement, each failing criterion's score increases by at least $\delta$ per iteration. Since scores are bounded in $[0, 1]$ and the maximum distance to threshold is $\tau$, each criterion requires at most $\lceil \tau / \delta \rceil$ iterations to reach $\tau$. With $n$ criteria, the worst case (sequential, non-overlapping improvement) yields $n \cdot \lceil \tau / \delta \rceil$ iterations. In practice, multiple criteria improve simultaneously, yielding substantially faster convergence.
\end{proof}

\textbf{Empirical validation}: The Focused Improvement assumption is supported by the self-refinement literature: \citet{lee2025refinebench} demonstrate that LLMs achieve near-perfect refinement when given targeted, structured feedback about specific quality dimensions. \efa{}'s per-criterion evaluation and emphasis weighting directly provides such structured feedback.

\subsection{Computational Complexity}
\label{sec:complexity}

Let $T$ denote the token cost of a single standard LLM generation call for prompt $P$.

\begin{itemize}[leftmargin=*]
    \item \textbf{Criteria generation}: 1 LLM call $\approx T_C$ tokens (typically $T_C \approx 0.5T$, as criteria are shorter than full responses).
    \item \textbf{Per iteration}: $n$ sub-steps of progressive generation ($n \cdot T$ tokens) + $n$ evaluation calls ($n \cdot T_E$ tokens, where $T_E \approx 0.3T$ per criterion).
    \item \textbf{Total over $K$ iterations}: $T_C + K \cdot n \cdot (T + T_E)$.
\end{itemize}

\textbf{Worst-case cost}: With $n = 5$ criteria, $K_{\max} = 3$ iterations, and $T_E = 0.3T$:
\begin{equation}
    T_{\text{EFA}} \approx 0.5T + 3 \cdot 5 \cdot (T + 0.3T) = 0.5T + 19.5T = 20T
\end{equation}

This represents a $20\times$ cost increase over single-pass generation. This is comparable to Best-of-20 sampling and lower than full multi-agent debate pipelines (which typically require $30{-}50T$). The cost is justified when output quality and reliability are prioritized over latency---common in enterprise, medical, legal, and high-stakes generation settings.

\textbf{Cost reduction strategies}: (a) \emph{Adaptive sub-stepping}: Skip sub-steps for criteria with high initial scores. (b) \emph{Early termination}: Exit the loop when all criteria pass after any sub-step, not only after the full iteration. (c) \emph{Batched evaluation}: Score all criteria in a single LLM call with multi-dimensional rubric prompting, reducing the $n \cdot T_E$ term to $\approx 1.5 T_E$.

\subsection{Complete Algorithm}

\begin{algorithm}[H]
\caption{Evaluation-First Attention (\efa{})}
\label{alg:efa}
\begin{algorithmic}[1]
\REQUIRE User prompt $P$, threshold $\tau$, max iterations $K_{\max}$, reattention strength $\alpha$, regression tolerance $\epsilon$
\ENSURE Response $R^*$
\STATE $C, \mathbf{w}^{(0)} \leftarrow \calC(P)$ \COMMENT{Generate criteria}
\STATE $R^{(0)} \leftarrow \emptyset$; \; $\text{locked} \leftarrow \emptyset$; \; $\mathbf{s}_{\text{best}} \leftarrow \mathbf{0}$
\FOR{$k = 1$ to $K_{\max}$}
    \STATE \COMMENT{\textbf{--- Criteria-Masked Progressive Generation ---}}
    \STATE $d_0^{(k)} \leftarrow R^{(k-1)}$
    \FOR{$j = 1$ to $|C|$}
        \STATE $M_j \leftarrow \text{CausalCriteriaMask}(j, |C|)$
        \STATE Construct prompt per Eq.~\ref{eq:prompt_construction} with priorities per Eq.~\ref{eq:priority}
        \STATE $d_j^{(k)} \leftarrow \calG(P, \; d_{j-1}^{(k)}, \; C \odot M_j, \; \hat{\mathbf{w}}^{(k-1)} \odot M_j)$
    \ENDFOR
    \STATE $R^{(k)} \leftarrow d_{|C|}^{(k)}$
    \STATE
    \STATE \COMMENT{\textbf{--- Per-Criterion Evaluation ---}}
    \STATE $\mathbf{s}^{(k)} \leftarrow \calE(R^{(k)}, C)$
    \STATE
    \STATE \COMMENT{\textbf{--- Criterion Checkpoint ---}}
    \FOR{$i = 1$ to $|C|$}
        \IF{$i \in \text{locked}$}
            \STATE $s_i^{(k)} \leftarrow s_{\text{best},i}$ \COMMENT{Retain locked score}
        \ELSIF{$s_{\text{best},i} \geq \tau$ \AND $s_i^{(k)} < s_{\text{best},i} - \epsilon$}
            \STATE $\text{locked} \leftarrow \text{locked} \cup \{i\}$ \COMMENT{Lock regressed criterion}
            \STATE $s_i^{(k)} \leftarrow s_{\text{best},i}$
        \ELSE
            \STATE $s_{\text{best},i} \leftarrow \max(s_{\text{best},i}, \; s_i^{(k)})$
        \ENDIF
    \ENDFOR
    \STATE
    \IF{$\min(\mathbf{s}^{(k)}) \geq \tau$}
        \RETURN $R^{(k)}$ \COMMENT{All criteria satisfied}
    \ENDIF
    \STATE
    \STATE \COMMENT{\textbf{--- Failure-Weighted Reattention ---}}
    \FOR{$i = 1$ to $|C|$}
        \IF{$i \notin \text{locked}$}
            \STATE $w_i^{(k)} \leftarrow w_i^{(k-1)} \cdot (1 + \alpha \cdot \max(0, \; \tau - s_i^{(k)}))$
        \ENDIF
    \ENDFOR
    \STATE $\hat{\mathbf{w}}^{(k)} \leftarrow \text{Normalize}(\mathbf{w}^{(k)})$
\ENDFOR
\RETURN $R^{(K_{\max})}$
\end{algorithmic}
\end{algorithm}

% ============================================================================
% 4. EXPERIMENTAL SETUP
% ============================================================================
\section{Experimental Design}
\label{sec:experiments}

We present our evaluation methodology below. Experimental results with quantitative comparisons will be reported in a forthcoming revision.\footnote{We release the architectural formalization and experimental design ahead of results to establish priority on the novel mechanisms (\cmpg{} and \fwrl{}) and to invite community feedback on the evaluation protocol.}

\subsection{Baselines}

We compare \efa{} against the following six baselines, designed to isolate the contribution of each architectural component:

\begin{enumerate}[leftmargin=*]
    \item \textbf{Single-pass}: Standard LLM generation without criteria or refinement. Measures the baseline quality achievable without any architectural intervention.
    \item \textbf{Self-Refine} \citep{madaan2023selfrefine}: Iterative holistic self-feedback and refinement. The primary comparison for iterative improvement without criteria structure.
    \item \textbf{Rubric-then-Score}: Generate criteria (same as \efa{} Step 1), generate response without criteria conditioning, then score with the rubric. Tests whether criteria-aware generation outperforms criteria-unaware generation followed by the same evaluation.
    \item \textbf{All-Criteria-at-Once}: Provide all criteria simultaneously to the generator in a single pass (no progressive masking, no iteration). Tests whether \cmpg{}'s progressive unmasking adds value over naive full-criteria prompting.
    \item \textbf{Uniform Reattention}: Generate with criteria, evaluate per-criterion, but re-generate with uniform weights on failure (no failure weighting). Tests whether \fwrl{}'s targeted reweighting outperforms uniform retry.
    \item \textbf{FusioN} \citep{agarwal2025fusion}: Multi-candidate synthesis without criteria guidance. Represents the state-of-the-art in inference-time quality scaling.
\end{enumerate}

\subsection{Ablation Studies}

To validate each component's individual contribution, we conduct four ablation conditions:
\begin{enumerate}[leftmargin=*]
    \item \textbf{$-$DynCriteria}: Replace dynamic per-query criteria with a fixed universal set of five criteria (relevance, accuracy, completeness, clarity, conciseness). Tests the value of dynamic criteria generation.
    \item \textbf{$-$\cmpg{}}: Remove progressive masking; present all criteria at once in every sub-step. Tests whether progressive unmasking improves over simultaneous conditioning.
    \item \textbf{$-$\fwrl{}}: Remove failure weighting; use uniform weights across all iterations. Tests whether targeted reweighting outperforms uniform iteration.
    \item \textbf{$-$Iteration}: Single generation pass with all criteria (no refinement loop). Tests whether iteration adds value beyond single-pass criteria-conditioned generation.
\end{enumerate}

\subsection{Evaluation Metrics}

\begin{itemize}[leftmargin=*]
    \item \textbf{Rubric Adherence Score (RAS)}: Mean per-criterion score across all criteria, averaged over test prompts. Range: $[0, 1]$.
    \item \textbf{All-Pass Rate (APR)}: Percentage of test prompts where all criteria meet threshold $\tau$. Measures reliability.
    \item \textbf{Iterations to Convergence (ITC)}: Mean number of iterations before all criteria pass or maximum reached. Measures efficiency.
    \item \textbf{Expected Calibration Error (ECE)}: Measures whether the system's aggregate score correlates with actual output quality as judged by humans. Binned across 10 confidence intervals.
    \item \textbf{Total Token Cost (TTC)}: Total tokens consumed across the full pipeline. Measures computational efficiency.
\end{itemize}

\subsection{Hyperparameter Settings}

Based on preliminary analysis, we propose the following default configuration: $n \in \{4, 5, 6\}$ criteria per query, threshold $\tau = 0.6$ (corresponding to rubric score $\geq 3/5$), maximum iterations $K_{\max} = 3$, reattention strength $\alpha \in \{1.0, 2.0, 5.0\}$ (studied via sensitivity analysis), and regression tolerance $\epsilon = 0.1$.

% ============================================================================
% 6. ANALYSIS AND DISCUSSION
% ============================================================================
\section{Analysis and Discussion}

\subsection{Why Evaluation-First Outperforms Evaluate-After}

The core insight supporting \efa{} comes from the self-refinement literature: structured, targeted feedback enables near-perfect refinement, while self-diagnosis is the bottleneck \citep{lee2025refinebench}. \efa{} eliminates the self-diagnosis problem entirely by pre-specifying what success looks like before generation begins. The model never needs to ``figure out what went wrong''---the per-criterion evaluation tells it exactly which dimensions failed and by how much.

\subsection{The Focal Property of Failure-Weighted Reattention}

\fwrl{} implements an inference-time analog of focal loss \citep{lin2017focal}. In object detection, focal loss down-weights well-classified examples to focus training on hard negatives. \fwrl{} analogously de-emphasizes (relatively) well-satisfied criteria to focus generation compute on failing dimensions. This creates an efficiency property: the system allocates more ``thinking budget'' to dimensions that need it most, rather than uniformly distributing effort across all quality dimensions regardless of their current satisfaction level.

\subsection{Criteria Masking as Quality Curriculum}

\cmpg{}'s progressive unmasking implements a form of curriculum learning \citep{bengio2009curriculum} at inference time. By satisfying fundamental criteria before introducing secondary ones, the generator builds a solid foundation that subsequent refinements can improve without undermining. This prevents the common failure mode of surface-level optimization at the expense of deep quality---analogous to how curriculum learning in training prevents models from memorizing surface patterns before learning fundamental structure.

\subsection{Connections to Test-Driven Development}

\efa{} draws a direct analogy to test-driven development (TDD) in software engineering: write the tests (criteria) before the code (response), then code (generate) to pass the tests (criteria). This analogy extends further: \cmpg{}'s progressive masking corresponds to TDD's practice of writing one test at a time and making it pass before adding the next, while \fwrl{}'s failure-weighted reattention corresponds to focusing debugging effort on failing tests rather than re-running the entire suite uniformly.

\subsection{Limitations}

\begin{itemize}[leftmargin=*]
    \item \textbf{Criteria quality dependence}: \efa{}'s effectiveness is bounded by the quality of generated criteria. Poor, overlapping, or irrelevant criteria degrade the entire pipeline. Criteria quality could be improved through the rubric refinement techniques of \citet{wang2026rrd}.
    \item \textbf{Compute cost}: With worst-case $20T$ token cost for $n=5$, $K=3$, \efa{} is not suitable for latency-sensitive applications. The cost reduction strategies in \S\ref{sec:complexity} partially address this, but a fundamental tradeoff between quality and cost remains.
    \item \textbf{Conflicting criteria}: When criteria inherently conflict (e.g., conciseness vs.\ completeness), failure-weighted reattention may oscillate between dimensions. The criterion checkpoint mechanism mitigates oscillation but may converge to a Pareto-suboptimal solution that satisfies each dimension at only the threshold level.
    \item \textbf{Evaluator reliability}: Per-criterion scoring is performed by an LLM judge, inheriting known biases of LLM-as-a-Judge approaches (position bias, verbosity bias, self-preference bias). Using a separate model for evaluation or calibrated evaluators could reduce this risk.
    \item \textbf{Prompt-level conditioning}: The current design operates through prompt engineering (Eq.~\ref{eq:prompt_construction}) rather than modifying model internals. While this maximizes applicability across API-accessible models, a deeper integration at the embedding or attention level could yield stronger conditioning effects.
\end{itemize}

% ============================================================================
% 7. CONCLUSION
% ============================================================================
\section{Conclusion}

We presented \textbf{Evaluation-First Attention} (\efa{}), an architecture that inverts the traditional generate-then-evaluate paradigm by generating evaluation criteria before generation and using them as structured conditioning targets. Our two novel mechanisms---\textbf{Criteria-Masked Progressive Generation} (\cmpg{}) and \textbf{Failure-Weighted Reattention} (\fwrl{})---have no direct prior art and together create a convergent, dimension-aware refinement loop. \efa{} transforms evaluation from a passive post-hoc measurement into an active generation specification, drawing a direct analogy to test-driven development in software engineering: define what ``good'' looks like, then build to satisfy it.

Future work includes: (a) deeper model integration via criteria embeddings in the attention mechanism, (b) adaptive criteria count selection based on prompt complexity, (c) multi-model configurations where different LLMs handle generation versus evaluation, and (d) domain-specific evaluation in code generation, medical question answering, and enterprise document synthesis.

% ============================================================================
% REFERENCES
% ============================================================================
\bibliographystyle{plainnat}

\begin{thebibliography}{99}

\bibitem[Madaan et al.(2023)]{madaan2023selfrefine}
Aman Madaan, Niket Tandon, Prakhar Gupta, Skyler Hallinan, Luyu Gao, Sarah Wiegreffe, Uri Alon, Nouha Dziri, Shrimai Prabhumoye, Yiming Yang, Shashank Gupta, Bodhisattwa Prasad Majumder, Katherine Hermann, Sean Welleck, Amir Yazdanbakhsh, and Peter Clark.
\newblock Self-Refine: Iterative Refinement with Self-Feedback.
\newblock In \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2023.

\bibitem[Shinn et al.(2023)]{shinn2023reflexion}
Noah Shinn, Federico Cassano, Ashwin Gopinath, Karthik Narasimhan, and Shunyu Yao.
\newblock Reflexion: Language Agents with Verbal Reinforcement Learning.
\newblock In \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2023.

\bibitem[Gou et al.(2024)]{gou2024critic}
Zhibin Gou, Zhihong Shao, Yeyun Gong, Yelong Shen, Yujiu Yang, Nan Duan, and Weizhu Chen.
\newblock CRITIC: Large Language Models Can Self-Correct with Tool-Interactive Critiquing.
\newblock In \emph{International Conference on Learning Representations (ICLR)}, 2024.

\bibitem[Liu et al.(2023)]{liu2023geval}
Yang Liu, Dan Iter, Yichong Xu, Shuohang Wang, Ruochen Xu, and Chenguang Zhu.
\newblock G-Eval: NLG Evaluation using GPT-4 with Better Human Alignment.
\newblock In \emph{Proceedings of the 2023 Conference on Empirical Methods in Natural Language Processing (EMNLP)}, 2023.

\bibitem[Jiang et al.(2023)]{jiang2023llmblender}
Dongfu Jiang, Xiang Ren, and Bill Yuchen Lin.
\newblock LLM-Blender: Ensembling Large Language Models with Pairwise Ranking and Generative Fusion.
\newblock In \emph{Proceedings of the 61st Annual Meeting of the Association for Computational Linguistics (ACL)}, 2023.

\bibitem[Lin et al.(2017)]{lin2017focal}
Tsung-Yi Lin, Priya Goyal, Ross Girshick, Kaiming He, and Piotr Doll\'{a}r.
\newblock Focal Loss for Dense Object Detection.
\newblock In \emph{IEEE International Conference on Computer Vision (ICCV)}, 2017.

\bibitem[Bengio et al.(2009)]{bengio2009curriculum}
Yoshua Bengio, J\'{e}r\^{o}me Louradour, Ronan Collobert, and Jason Weston.
\newblock Curriculum Learning.
\newblock In \emph{International Conference on Machine Learning (ICML)}, 2009.

\bibitem[Bai et al.(2022)]{bai2022constitutional}
Yuntao Bai, Saurav Kadavath, Sandipan Kundu, Amanda Askell, Jackson Kernion, Andy Jones, Anna Chen, Anna Goldie, Azalia Mirhoseini, Cameron McKinnon, et al.
\newblock Constitutional AI: Harmlessness from AI Feedback.
\newblock \emph{arXiv preprint arXiv:2212.08073}, 2022.

\bibitem[Yang and Klein(2021)]{yang2021fudge}
Kevin Yang and Dan Klein.
\newblock FUDGE: Controlled Text Generation With Future Discriminators.
\newblock In \emph{Proceedings of the 2021 Conference of the North American Chapter of the Association for Computational Linguistics (NAACL)}, 2021.

\bibitem[Dathathri et al.(2020)]{dathathri2020plug}
Sumanth Dathathri, Andrea Madotto, Janice Lan, Jane Hung, Eric Frank, Piero Molino, Jason Yosinski, and Rosanne Liu.
\newblock Plug and Play Language Models: A Simple Approach to Controlled Text Generation.
\newblock In \emph{International Conference on Learning Representations (ICLR)}, 2020.

\bibitem[Ho and Salimans(2022)]{ho2022classifier}
Jonathan Ho and Tim Salimans.
\newblock Classifier-Free Diffusion Guidance.
\newblock In \emph{NeurIPS 2021 Workshop on Deep Generative Models and Downstream Applications}, 2022.

\bibitem[Chen et al.(2025)]{chen2025magicore}
Justin Chih-Yao Chen, Archiki Prasad, Swarnadeep Saha, and Mohit Bansal.
\newblock MAgICoRe: Multi-Agent, Iterative, Coarse-to-Fine Refinement for Reasoning.
\newblock In \emph{Proceedings of the 2025 Conference on Empirical Methods in Natural Language Processing (EMNLP)}, 2025.

\bibitem[Lee et al.(2025)]{lee2025refinebench}
Hyunseok Lee, Seunghyuk Oh, Jaehyung Kim, Jinwoo Shin, and Jihoon Tack.
\newblock RefineBench: Evaluating Self-Refinement Capabilities of Large Language Models.
\newblock \emph{arXiv preprint arXiv:2411.00548}, 2025.

\bibitem[Agarwal et al.(2025)]{agarwal2025fusion}
Rishabh Agarwal, Nino Vieillard, Yongchao Zhou, Avi Singh, and Olivier Bachem.
\newblock Making, not Taking, the Best of N.
\newblock \emph{arXiv preprint arXiv:2510.00931}, 2025.

\bibitem[Li et al.(2025)]{li2025openrubrics}
Tianyi Li, Tong Xiao, Zhixian Zhang, et al.
\newblock OpenRubrics: Towards Scalable Synthetic Rubric Generation for Reward Modeling and LLM Alignment.
\newblock \emph{arXiv preprint arXiv:2510.07743}, 2025.

\bibitem[Zhang et al.(2025)]{zhang2025carmo}
Zhiyu Zhang, Shujian Huang, Xue Han, and Jiajun Chen.
\newblock CARMO: Dynamic Criteria Generation for Context-Aware Reward Modelling.
\newblock In \emph{Findings of the Association for Computational Linguistics: ACL 2025}, 2025.

\bibitem[Wang et al.(2026)]{wang2026rrd}
Yufei Wang, Jiayi Pan, Yaswanth Narsupalli, Tong Zhang, and Yulia Tsvetkov.
\newblock Rethinking Rubric Generation for Improving LLM Judge and Reward Modeling for Open-ended Tasks.
\newblock \emph{arXiv preprint arXiv:2602.05125}, 2026.

\bibitem[Chen et al.(2025b)]{chen2025autoscore}
Mengxiao Chen, Yunfei Wang, Yanqing Wang, et al.
\newblock AutoSCORE: Enhancing Automated Scoring with Multi-Agent Large Language Models via Structured Component Recognition.
\newblock \emph{arXiv preprint arXiv:2509.21910}, 2025.

\bibitem[Ravi et al.(2025)]{ravi2025llmloop}
Karandeep Ravi, Valerio Terragni, and Gunel Jahangirova.
\newblock LLMLOOP: Improving LLM-Generated Code and Tests via Iterative Feedback.
\newblock In \emph{IEEE International Conference on Software Maintenance and Evolution (ICSME)}, 2025.

\end{thebibliography}

% ============================================================================
% APPENDIX
% ============================================================================
\appendix
\section{Prompt Templates}
\label{app:prompts}

We provide the core prompt templates used in each component of \efa{} to facilitate reproducibility.

\subsection{Criteria Generator Prompt}

\begin{small}
\begin{verbatim}
You are an evaluation criteria designer. Given a user 
prompt, generate {n} non-overlapping evaluation criteria 
ordered from most fundamental to least fundamental.

For each criterion, provide:
1. Name (2-4 words)
2. Definition (1 sentence)
3. Scoring rubric:
   - Score 1: [description of lowest quality]
   - Score 2: [description of below-average quality]
   - Score 3: [description of acceptable quality]
   - Score 4: [description of good quality]
   - Score 5: [description of excellent quality]

Requirements:
- Criteria must be measurable and non-overlapping
- Order by fundamentality: correctness before style
- Each criterion must be independently scorable
- Criteria must be specific to this prompt, not generic

User prompt: {P}
\end{verbatim}
\end{small}

\subsection{Progressive Generation Prompt (Sub-step $j$)}

\begin{small}
\begin{verbatim}
Generate a response to the user prompt below.

{If draft exists: "Improve the following draft:
{draft}"}

Your response MUST satisfy the following criteria,
listed in priority order:

{For each visible criterion c_i where M_j[i]=1:}
[{priority_label}] {criterion_name}:
  {criterion_definition}

User prompt: {P}
\end{verbatim}
\end{small}

\subsection{Per-Criterion Evaluation Prompt}

\begin{small}
\begin{verbatim}
Evaluate the following response against ONE specific 
criterion. Score ONLY this criterion on a 1-5 scale 
using the rubric provided.

Criterion: {criterion_name}
Definition: {criterion_definition}
Rubric:
  1: {rubric_1}
  2: {rubric_2}
  3: {rubric_3}
  4: {rubric_4}
  5: {rubric_5}

User prompt: {P}
Response: {R}

Provide your score (1-5) and a one-sentence 
justification. Output format:
Score: [1-5]
Reason: [one sentence]
\end{verbatim}
\end{small}

\section{Worked Example}
\label{app:example}

We illustrate \efa{} on a sample prompt to clarify the mechanism.

\textbf{User prompt}: ``Explain how photosynthesis works for a 10-year-old.''

\textbf{Step 1 --- Criteria generated}:
\begin{enumerate}
    \item \emph{Scientific Accuracy} (fundamental): Facts about photosynthesis must be correct.
    \item \emph{Age Appropriateness}: Language and concepts suitable for a 10-year-old.
    \item \emph{Completeness}: Covers inputs (sunlight, water, CO$_2$), process, and outputs (glucose, O$_2$).
    \item \emph{Engagement}: Uses analogies, examples, or vivid language to maintain interest.
\end{enumerate}

\textbf{Step 2 --- Progressive generation} (iteration $k=1$):
\begin{itemize}
    \item Sub-step 1: Generate attending to \{Scientific Accuracy\} only $\rightarrow$ draft focuses on correct facts.
    \item Sub-step 2: Refine attending to \{Scientific Accuracy, Age Appropriateness\} $\rightarrow$ simplifies vocabulary.
    \item Sub-step 3: Refine attending to \{$c_1, c_2$, Completeness\} $\rightarrow$ adds missing components.
    \item Sub-step 4: Refine attending to all four criteria $\rightarrow$ adds analogies for engagement.
\end{itemize}

\textbf{Step 3 --- Evaluation}: Scores $= (5, 4, 3, 2)$. With $\tau = 0.6$ (i.e., score $\geq 3$), criteria 1--3 pass but criterion 4 (Engagement, score 2/5 $= 0.4$) fails.

\textbf{Step 4 --- Reattention}: With $\alpha = 2.0$, Engagement weight boosted: $w_4^{(1)} = 0.25 \cdot (1 + 2.0 \cdot (0.6 - 0.4)) = 0.35$. After renormalization, Engagement receives \texttt{[HIGH]} priority in iteration $k=2$.

\textbf{Step 5 --- Iteration $k=2$}: Progressive generation repeats with boosted Engagement emphasis. Re-evaluation yields scores $= (5, 4, 4, 4)$. All criteria pass. Output returned.

\end{document}
